The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic

### 2021-05-21 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
First version for submission to PLoS ONE

The emergence of the novel SARS-CoV-2 coronavirus in 2019, and the
global pandemic that followed in its wake, led to an explosive growth of
research. According to Fraser et al. {[}1{]}, over 125,000
COVID-19-related papers were released in the first ten months from the
first confirmed case of the disease. Of these, more than 30,000 were
shared in pre-print servers, the use of which also exploded in the past
year {[}2--4{]}.

@@

In the context of COVID-19, a topic that has grabbed the imagination of
numerous thinkers has been the prospect of life in cities after the
pandemic {[}18{]}. The fact that the worst of the pandemic was initially
felt in dense population centers such as Wuhan, Milan, Madrid, and New
York, brought a torrent of research into the associations between
density and the spread of the pandemic. Some important questions hang on
the results of these research efforts. For example, are lower density
regions safer from the pandemic? Are de-densification policies
warranted, at least in the short term? And in the longer term, will the
risks of life in high density regions presage a flight from cities? Over
the past year, numerous papers have sought to throw light into the
underlying issue of density and the pandemic; nonetheless the results,
as will be detailed next, remain mixed. Further, to complicate matters,
precious few of these studies appear to be sufficiently open to support
independent verification.

@@

\hypertarget{background}{%
\section{Background}\label{background}}

The concern with population density and the spread of the virus during
the COVID-19 pandemic was fueled, at least in part, by dramatic scenes
seen in real-time around the world from large urban centers such as
Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to
believe that higher density may have a positive association with the
transmission of a contagious virus. It has long been known that the
potential for inter-personal contact is greater in regions with higher
density {[}see for example the research on urban fields and
time-geography 20,21,22{]}. Mathematically, models of exposure and
contagion indicate that higher densities can catalyze the transmission
of contagious diseases {[}23,24{]}. Models such as these were likely at
the root of messages, by some figures in positions of authority, that
regions with sparse population densities faced lower risks from the
pandemic\footnote{Governor Kristi Noem of South Dakota, for example,
  claimed that sparse population density allowed her state to face the
  pandemic down without the need for strict policy interventions
  \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}}.

As Rocklöv and Sjödin {[}23{]} note, however, mathematical models of
contagion are valid at small-to-medium spatial scales (and presumably,
small temporal scales too, such as time spent in restaurants, concert
halls, cruises), and the results do not necessarily transfer to larger
spatial units and different time scales. There are solid reasons for
this: while in a restaurant, one can hardly avoid being in proximity to
other customers-however, a person can choose to (or be forced to as a
matter of policy) not go to a restaurant in the first place.
Nonetheless, the idea that high density correlates with high
transmission is so intuitive that it is often taken for granted even at
larger scales {[}e.g., 25,26{]}. At larger scales, however, there exists
the possibility of behavioral adaptations, which are difficult to
capture in the mechanistic framework of differential equations {[}or can
be missing in agent-based models, e.g., 27{]}; these adaptations, in
fact, can be a key aspect of disease transmission.

@@

In this respect, the literature to date remains inconclusive.

@@

As can be seen, these studies are implemented at different scales in
different regions of the world. They also use a range of techniques,
from correlation analysis, to multivariate regression, spatial
regressions, and machine learning techniques. This is natural and to be
expected: individual researchers have only limited time and expertise.
This is why reproducibility is important. To pick an example (which will
be further elaborated in the following sections), the study of Sy et
al.~{[}{[}19{]}; hereafter SWN{]} would immediately grab the attention
of a researcher with expertise in spatial modelling.

SWN investigated the basic reproductive number of COVID-19 in US
counties, and its association with population density, median household
income, and prevalence of private mobility. The basic reproductive
number can only be calculated reliably with a minimum number of cases,
and a large number of counties did not meet such threshold. As
researchers do, SWN made certain seemingly reasonable modelling
decision, for instance basing their analysis only on counties with valid
observations. An experiences spatial modeler would likely ask some of
the following questions on reading SWN's paper: how were missing
counties treated? Is it possible to spatially interpolate the missing
observations? What are the implications of the spatial sampling
framework used in the analysis? And, was there evidence of spatial
autocorrelation in the residuals of the models? These are questions that
in most cases would not occur to a researcher who has not been exposed
to spatial statistics or spatial econometrics. Nonetheless, they are
relevant and important. Fortunately, SWN give an example of a reasonably
open, reproducible research product: their paper is accompanied by (most
of) the data and (most of) the code used in the analysis. This means
that an independent expert can, with only a moderate investment of time
and effort, replicate the results in the paper, as well as ask
additional questions.

@@

In the following sections, the analysis of RWN is replicated, some
relevant questions from the perspective of a spatial modeler are asked,
and the data are reanalyzed.

@@

SWN examined the association between the basic reproductive number of
COVID-19 and population density. The basic reproductive number \(R_0\)
is a summary measure of contact rates, probability of transmission of a
pathogen, and duration of infectiousness. In rough terms, \(R_0\)
measures how many new infections each infections begets. Infectious
disease outbreaks generally tend to die out when \(R_0<1\), and to grow
when \(R_0>1\). Reliable calculation of \(R_0\) requires a minimum
number of cases to be able to assume that there is community
transmission of the pathogen. Accordingly, SNW based their analysis only
on counties that had at least 25 cases or more at the end of the
exponential growth phase (see Fig. \ref{fig:R0-map}). Their final sample
included 1,151 counties in the US, including in Alaska, Hawaii, Puerto
Rico, and island territories.

@@

SWN used for their multivariate analysis mixed linear models. This is a
reasonable modelling choice: \(R_0\) is an interval-ratio variable that
is suitably modeled using linear regression; further, as SWN note there
is a likelihood that the process in not independent ``among counties
within each state, potentially due to variable resource allocation and
differing health systems across states'' (p.~3). A mixed linear model
accounts for this by introducing random components (in the case of SNW,
random intercepts at the state level). SWN estimated various models with
different combinations of variables, including median household income
and prevalence of travel by private transportation. These are sensible
controls, given potential variations in behavior: people in more
affluent counties may have greater opportunities to work from home, and
use of private transportation reduces contact with strangers. Moreover,
they also conducted various sensitivity analyses. After these efforts,
SNW conclude that there is a positive association between the basic
reproductive number and population density at the level of counties in
the US.

Table \ref{tab:swn-results} replicates the first three models of SWN
(the fourth model did not have any significant variables; see Table 1 in
SNW). It is possible to verify that the results match with the minor
(and irrelevant) exception of the magnitude of the coefficient for
travel by private transportation, which is due to a difference in the
input (here the variable is one percent units, whereas in SWN it was ten
percent units). The intercept in these models is a random variable, and
the standard deviation is reported in the fourth row of Table
\ref{tab:swn-results}. It is useful to map the random intercepts: as
seen in Figure \ref{fig:random-terms-map}, other things being equal,
counties in Texas tended to have somewhat lower values of \(R_0\) (i.e.,
a negative random intercept), whereas counties in South Dakota tended to
have higher values of \(R_0\). The key of the analysis, that population
density has a positive association with the basic reproductive number,
is verified. But is it really?

@@

\caption{\label{tab:tabulate-swn-results}\label{tab:swn-results}Replicating SWN: Models 1-3}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllllrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Model 1} & \multicolumn{2}{c}{Model 2} & \multicolumn{2}{c}{Model 3} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
Variable & beta & 95\% CI & beta & 95\% CI & beta & 95\% CI\\
\midrule
Intercept & 2.274 & [2.167, 2.381] & 3.347 & [2.676, 4.018] & 3.386 & [2.614, 4.157]\\
Log of population density & 0.162 & [0.133, 0.191] & 0.145 & [0.115, 0.176] & 0.147 & [0.113, 0.18]\\
Percent of private transportation &  &  & -0.013 & [-0.02, -0.005] & -0.013 & [-0.021, -0.005]\\
Median household income (\$10,000) &  &  &  &  & -0.003 & [-0.033, 0.026]\\
Standard deviation (Intercept) & 0.166 & [0.108, 0.254] & 0.136 & [0.081, 0.229] & 0.137 & [0.081, 0.232]\\
\addlinespace
Within-group standard error & 0.665 & [0.638, 0.693] & 0.665 & [0.638, 0.693] & 0.665 & [0.638, 0.694]\\
\bottomrule
\end{tabular}}
\end{table}

@@

Thanks to the availability of code and data, it was possible to verify
the results reported by SWN. As noted earlier, though, an experienced
spatial modeller would have asked themselves about the implications of
the spatial sampling procedure used by SWN. The decision to use a sample
of counties with reliable basic reproductive numbers, although seemingly
sensible, results in a non-random sampling scheme. Turning our attention
back to Figure \ref{fig:R0-map}, we see that many counties without
reliable values of \(R_0\) are in more rural, less dense parts of the
United States. The fact that \(R_0\) could not be accurately computed in
those counties does not mean that there was no transmission of the
virus: it simply means that we do not know with precision whether that
was the case. The low number of cases may be related to low population
and/or low population density. This is intriguing, to say the least: by
excluding cases in a non-random way such as this, we are in effect
\emph{truncating} the sample.

The issue with sample truncation is that parameter estimates can become
unreliable.

Alas, despite decades' worth of developments in the field of
geographical analysis, not all research presented to date has used
proper methods for the study of COVID-19. While the answer to the
question ``do spatial effects really matter in regression analysis'' was
definitively answered in the positive at least 40 years ago, in practice
many researchers continue to ignore the pitfalls of ignoring them. In
this paper, I present a reanalysis of the data used by Sy et al.~(2021)
to study the correlations between the basic reproductive number of
COVID-19 and population density in counties. I highlight two related
issues: non-systematic sampling in space and spatial autocorrelation.
The reanalysis is based on the use of tobit models to account for
non-systematic sampling, and spatially autoregressive tobit models to
account for spatial autocorrelation in the data generation process. The
reanalysis highlights the importance of openness and reproducibility in
COVID-19 research. Finally, the results provide a sobering example of
the risks of not using appropriate methods in the analysis of
geographical data. Furthermore,

\hypertarget{fit-tobit-version-of-models}{%
\subsection{Fit tobit version of
models}\label{fit-tobit-version-of-models}}

@@

\caption{\label{tab:unnamed-chunk-3}\label{tab:tobit-results}Marginal effects of sample selection models}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllllrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Tobit Model 1} & \multicolumn{2}{c}{Tobit Model 2} & \multicolumn{2}{c}{Tobit Model 3} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
Variable & m.e. & 95\% CI & m.e. & 95\% CI & m.e. & 95\% CI\\
\midrule
Log of population density & 0.264 & [0.246, 0.281] & 0.264 & [0.247, 0.281] & 0.233 & [0.216, 0.25]\\
Percent of private transportation &  &  & 0.006 & [0.002, 0.01] & 0.014 & [0.01, 0.019]\\
Median household income (\$10,000) &  &  &  &  & 0.105 & [0.087, 0.122]\\
\bottomrule
\end{tabular}}
\end{table}

\hypertarget{spatially-autoregressive-tobit}{%
\subsection{Spatially autoregressive
tobit}\label{spatially-autoregressive-tobit}}

Fit spatially autoregressive tobit:

@@

Words go here.

@@

>>>>

The emergence of the novel SARS-CoV-2 coronavirus in 2019, and the
global pandemic that followed in its wake, led to an explosive growth of
research around the globe. According to Fraser et al. {[}1{]}, over
125,000 COVID-19-related papers were released in the first ten months
from the first confirmed case of the disease. Of these, more than 30,000
were shared in pre-print servers, the use of which also exploded in the
past year {[}2--4{]}.

@@

In the context of COVID-19, a topic that has grabbed the imagination of
numerous thinkers has been the prospect of life in cities after the
pandemic {[}18{]}. The fact that the worst of the pandemic was initially
felt in dense population centers such as Wuhan, Milan, Madrid, and New
York, brought a torrent of research into the associations between
density and the spread of the pandemic. Some important questions hang on
the results of these research efforts. For example, are lower density
regions safer from the pandemic? Are de-densification policies
warranted, even if just in the short term? And in the longer term, will
the risks of life in high density regions presage a flight from cities?
Over the past year, numerous papers have sought to throw light into the
underlying issue of density and the pandemic; nonetheless the results,
as will be detailed next, remain mixed. Further, to complicate matters,
precious few of these studies appear to be sufficiently open to support
independent verification.

@@

\hypertarget{background-the-intuitive-relationship-between-density-and-spread-of-contagious-diseases}{%
\section{Background: the intuitive relationship between density and
spread of contagious
diseases}\label{background-the-intuitive-relationship-between-density-and-spread-of-contagious-diseases}}

The concern with population density and the spread of the virus during
the COVID-19 pandemic was fueled, at least in part, by dramatic scenes
seen in real-time around the world from large urban centers such as
Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to
believe that higher density may have a positive association with the
transmission of a contagious virus. It has long been known that the
potential for inter-personal contact is greater in regions with higher
density {[}see for example the research on urban fields and
time-geography 20,21,22{]}. Mathematically, models of exposure and
contagion indicate that higher densities can catalyze the transmission
of contagious diseases {[}23,24{]}. The idea is intuitive and likely at
the root of messages, by some figures in positions of authority, that
regions with sparse population densities faced lower risks from the
pandemic\footnote{Governor Kristi Noem of South Dakota, for example,
  claimed that sparse population density allowed her state to face the
  pandemic down without the need for strict policy interventions
  \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}}.

As Rocklöv and Sjödin {[}23{]} note, however, mathematical models of
contagion are valid at small-to-medium spatial scales (and presumably,
small temporal scales too, such as time spent in restaurants, concert
halls, cruises), and the results do not necessarily transfer to larger
spatial units and different time scales. There are solid reasons for
this: while in a restaurant, one can hardly avoid being in proximity to
other customers-however, a person can choose to (or be forced to as a
matter of policy) not go to a restaurant in the first place.
Nonetheless, the idea that high density correlates with high
transmission is so seemingly reasonable that it is often taken for
granted even at larger scales {[}e.g., 25,26{]}. At larger scales,
however, there exists the possibility of behavioral adaptations, which
are difficult to capture in the mechanistic framework of differential
equations {[}or can be missing in agent-based models, e.g., 27{]}; these
adaptations, in fact, can be a key aspect of disease transmission.

@@

\hypertarget{background-but-what-does-the-literature-say}{%
\section{Background: but what does the literature
say?}\label{background-but-what-does-the-literature-say}}

When it comes to population density and the spread of COVID-19, the
international literature to date remains inconclusive.

@@

As can be seen, these studies are implemented at different scales in
different regions of the world. They also use a range of techniques,
from correlation analysis, to multivariate regression, spatial
regressions, and machine learning techniques. This is natural and to be
expected: individual researchers have only limited time and expertise.
This is why reproducibility is important. To pick an example (which will
be further elaborated in later sections of this paper), the study of Sy
et al.~{[}{[}19{]}; hereafter SWN{]} would immediately grab the
attention of a researcher with a somewhat different toolbox.

\hypertarget{reproducibility-of-research}{%
\section{Reproducibility of
research}\label{reproducibility-of-research}}

SWN investigated the basic reproductive number of COVID-19 in US
counties, and its association with population density, median household
income, and prevalence of private mobility. For their multivariate
analysis, SWN used mixed linear models. This is a reasonable modelling
choice: \(R_0\) is an interval-ratio variable that is suitably modeled
using linear regression; further, as SWN note there is a likelihood that
the process in not independent ``among counties within each state,
potentially due to variable resource allocation and differing health
systems across states'' (p.~3). A mixed linear model accounts for this
by introducing random components (in the case of SWN, random intercepts
at the state level). SWN estimated various models with different
combinations of variables, including median household income and
prevalence of travel by private transportation. These are sensible
controls, given potential variations in behavior: people in more
affluent counties may have greater opportunities to work from home, and
use of private transportation reduces contact with strangers. Moreover,
they also conducted various sensitivity analyses. After these efforts,
SWN conclude that there is a positive association between the basic
reproductive number and population density at the level of counties in
the US.

One salient aspect of the analysis in SWN is that the basic reproductive
number can only be calculated reliably with a minimum number of cases,
and a large number of counties did not meet such threshold. As
researchers do, SWN made modelling decisions, in this case basing their
analysis only on counties with valid observations. A modeler with
expertise in different methods would likely ask some of the following
questions on reading SWN's paper: how were missing counties treated?
What are the implications of the spatial sampling framework used in the
analysis? Is it possible to spatially interpolate the missing
observations? And, was there evidence of spatial autocorrelation in the
residuals of the models? These questions are relevant and their
implications important. Fortunately, SWN are an example of a reasonably
open, reproducible research product: their paper is accompanied by (most
of) the data and (most of) the code used in the analysis. This means
that an independent expert can, with only a moderate investment of time
and effort, reproduce the results in the paper, as well as ask
additional questions.

@@

In the following sections, the analysis of RWN is reproduced, some
relevant questions from the perspective of a spatial modeler are asked,
and the data are reanalyzed.

@@

SWN examined the association between the basic reproductive number of
COVID-19 and population density. The basic reproductive number \(R_0\)
is a summary measure of contact rates, probability of transmission of a
pathogen, and duration of infectiousness. In rough terms, \(R_0\)
measures how many new infections each infections begets. Infectious
disease outbreaks generally tend to die out when \(R_0<1\), and to grow
when \(R_0>1\). Reliable calculation of \(R_0\) requires a minimum
number of cases to be able to assume that there is community
transmission of the pathogen. Accordingly, SWN based their analysis only
on counties that had at least 25 cases or more at the end of the
exponential growth phase (see Fig. \ref{fig:R0-map}). Their final sample
included 1,151 counties in the US, including in Alaska, Hawaii, Puerto
Rico, and island territories.

@@

Table \ref{tab:swn-results} reproduces the first three models of SWN
(the fourth model did not have any significant variables; see Table 1 in
SWN). It is possible to verify that the results match, with only the
minor (and irrelevant) exception of the magnitude of the coefficient for
travel by private transportation, which is due to a difference in the
input (here the variable is one percent units, whereas in SWN it was ten
percent units). The mixed linear model gives random intercepts (i.e.,
the intercept is a random variable), and the standard deviation is
reported in the fourth row of Table \ref{tab:swn-results}. It is useful
to map the random intercepts: as seen in Figure
\ref{fig:random-terms-map}, other things being equal, counties in Texas
tend to have somewhat lower values of \(R_0\) (i.e., a negative random
intercept), whereas counties in South Dakota tend to have higher values
of \(R_0\). The key of the analysis, after extensive sensitivity
analysis, is a robust finding that population density has a positive
association with the basic reproductive number. But does it?

@@

\caption{\label{tab:tabulate-swn-results}\label{tab:swn-results}Reproducing SWN: Models 1-3}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllllrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Model 1} & \multicolumn{2}{c}{Model 2} & \multicolumn{2}{c}{Model 3} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
Variable & beta & 95\% CI & beta & 95\% CI & beta & 95\% CI\\
\midrule
Intercept & 2.274 & [2.167, 2.381] & 3.347 & [2.676, 4.018] & 3.386 & [2.614, 4.157]\\
Log of population density & 0.162 & [0.133, 0.191] & 0.145 & [0.115, 0.176] & 0.147 & [0.113, 0.18]\\
Percent of private transportation &  &  & -0.013 & [-0.02, -0.005] & -0.013 & [-0.021, -0.005]\\
Median household income (\$10,000) &  &  &  &  & -0.003 & [-0.033, 0.026]\\
Standard deviation (Intercept) & 0.166 & [0.108, 0.254] & 0.136 & [0.081, 0.229] & 0.137 & [0.081, 0.232]\\
\addlinespace
Within-group standard error & 0.665 & [0.638, 0.693] & 0.665 & [0.638, 0.693] & 0.665 & [0.638, 0.694]\\
\bottomrule
\end{tabular}}
\end{table}

@@

The preceding section shows that thanks to the availability of code and
data, it is possible to verify the results reported by SWN. As noted
earlier, though, an independent researcher might have wondered about the
implications of the spatial sampling procedure used by SWN. The decision
to use a sample of counties with reliable basic reproductive numbers,
although apparently sensible, results in a non-random spatial sampling
scheme. Turning our attention back to Figure \ref{fig:R0-map}, we form
the impression that many counties without reliable values of \(R_0\) are
in more rural, less dense parts of the United States. This impression is
reinforced when we overlay the boundaries of urban areas with population
greater than 50,000 on the counties with valid values of \(R_0\) (see
Figure \ref{fig:urban-areas-map}). The fact that \(R_0\) could not be
accurately computed in many counties without large urban areas does not
mean that there was no transmission of the virus: it simply means that
we do not know with precision whether that was the case. The low number
of cases may be related to low population and/or low population density.
This is intriguing, to say the least: by excluding cases based on the
ability to calculate \(R_0\) we are potentially \emph{censoring} the
sample in a non-random way.

\begin{figure}
\includegraphics[width=1\linewidth]{R0-Density-Reanalysis_files/figure-latex/urban-areas-map-1} \caption{\label{fig:urban-areas-map}Urban areas with population > 50,000 (Alaska, Hawaii, Puerto Rico, and territories not shown).}\label{fig:urban-areas-map}
\end{figure}

A problematic issue with non-random sample selection is that parameter
estimates can become unreliable, and numerous techniques have been
developed over time to address this. A model useful for sample selection
problems is Heckman's selection model {[}see 62{]}. The selection model
is in fact a system of two equations, as follows:

\[
\begin{array}{c}
y_i^{S*} = \beta^{S\prime}x_i^S+\epsilon_i^S\\
y_i^{O*} = \beta^{O\prime}x_i^O+\epsilon_i^O
\end{array}
\] \noindent where \(y_i^{S*}\) is a latent variable for the sample
selection process, and \(y_i^{O*}\) is the latent outcome. Vectors
\(x_i^S\) and \(x_i^O\) are explanatory variables (with the possibility
that \(x_i^S = x_i^S\)). Both equations include random terms (i.e.,
\(\epsilon_i^S\) and \(\epsilon_i^O\)) The first equation is designed to
model the \emph{probability} of sampling, and the second equation the
outcome of interest (say \(R_0\)). The random terms are jointly
distributed and correlated with parameter \(\rho\).

What the analyst observes is the following: \[
y_i^S =
\begin{cases}
0 & \text{if } y_i^{S*} < 0\\
1 & \text{otherwise}
\end{cases}
\] \noindent and: \[
y_i^O =
\begin{cases}
0 & \text{if } y_i^{S} = 0\\
y_i^{O*} & \text{otherwise}
\end{cases}
\]

In other words, the outcome of interest is observed \emph{only} for
certain cases (\(y_i^S=1\), i.e., for sampled observations). The
probability of sampling depends on \(x_i^S\). For the cases observed,
the outcome \(y_i^O\) depends on \(x_i^O\).

A sample selection model is estimated using the same selection of
variables as SWN Model 3. This is Sample Selection Model 1 in Table
\ref{tab:selection-results}. The first thing to notice about this model
is that the sample selection process and the outcome are not independent
(\(\rho\ne0\) with 5\% of confidence). The selection equation indicates
that the probability of a county to be in the sample increases with
population density (but at a decreasing rate due to the
log-transformation), when travel by private modes is more prevalent, and
as median household income in the county is higher. This is in line with
the impression left by Figure \ref{fig:urban-areas-map} that counties
with reliable values of \(R_0\) tended to be those with larger urban
centers. Once that the selection probabilities are accounted for in the
model, several things happen with the outcomes model. First, the
coefficient for population density is still positive, but the magnitude
changes: in effect, it appears that the effect of density is more
pronounced than what SWN Model 3 indicated. The coefficient for percent
of private transportation changes signs. And the coefficient for median
household income is now significant.

The second model in Table \ref{tab:selection-results} (Selection Model
2) changes the way the variables are entered into the model. The
log-transformation of density in SWN and Selection Model 1 assumes that
the association between density and \(R_0\) is monotonically increasing
(if the sign of the coefficient is positive) or decreasing (if the sign
of the coefficient is negative). There are some indications that the
relationship may actually not be monotonical. For example, Paez et al.
{[}32{]} found a positive (if non-significant) relationship between
density and incidence of COVID-19 in the provinces of Spain at the
beginning of the pandemic. This changed to a negative (and significant)
relationship during the lockdown. In the case of the US, Fielding-Miller
et al. {[}44{]} found that the association between COVID-19 deaths and
population density was positive in rural counties, but negative in urban
counties. A variable transformation that allows for non-monotonic
changes in the relationship is the square of the density.

As seen in the table, Selection Model 2 replaces the log-transformation
of population density with a quadratic expansion. The results of this
analysis indicate that with this variable transformation, the selection
and outcome processes are still not independent (\(\rho\ne0\) with 5\%
of confidence). But a few interesting things emerge. When we examine the
outcomes model, we see that the quadratic expansion has a positive
coefficient for the first order term, but a negative coefficient for the
second order term. This indicates that \(R_0\) initially tends to
increase with higher density, but only up to a point, after which the
negative second term (which grows more rapidly due to the square),
becomes increasingly dominant. Secondly, the sign of the coefficient for
travel by private transportation becomes negative again. This, of
course, makes more sense than the positive sign of Selection Model 1: if
people tend to travel in private transportation, the potential for
contact should be lower instead of higher. And finally median household
income is no longer significant.

@@

\caption{\label{tab:tabulate-sample-selection-results}\label{tab:selection-results}Estimation results of sample selection models}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lcccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Selection Model 1} & \multicolumn{2}{c}{Selection Model 2} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Variable & $\beta$ & 95\% CI & $\beta$ & 95\% CI\\
\midrule
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Sample Selection Model}}\\
\hspace{1em}Intercept & -2.237 & [-3.109, -1.365] & -7.339 & [-8.381, -6.297]\\
\hspace{1em}Log of population density & 0.385 & [0.352, 0.418] &  & \\
\hspace{1em}Density (1,000 per sq.km) &  &  & 2.484 & [2.13, 2.838]\\
\hspace{1em}Density squared &  &  & -0.387 & [-0.473, -0.3]\\
\hspace{1em}Percent of private transportation & 0.025 & [0.016, 0.034] & 0.057 & [0.046, 0.067]\\
\hspace{1em}Median household income (10,000) & 0.202 & [0.168, 0.235] & 0.32 & [0.283, 0.357]\\
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Outcome Model}}\\
\hspace{1em}Intercept & 0.605 & [-0.257, 1.466] & 2.784 & [1.652, 3.915]\\
\hspace{1em}Log of population density & 0.39 & [0.354, 0.426] &  & \\
\hspace{1em}Density (1,000 per sq.km) &  &  & 0.758 & [0.509, 1.008]\\
\hspace{1em}Density squared &  &  & -0.132 & [-0.187, -0.077]\\
\hspace{1em}Percent of private transportation & 0.01 & [0.001, 0.018] & -0.011 & [-0.021, -0.001]\\
\hspace{1em}Median household income (\$10,000) & 0.126 & [0.094, 0.159] & 0.002 & [-0.033, 0.037]\\
$\sigma$ & 0.954 & [0.904, 1.003] & 0.684 & [0.652, 0.716]\\
$\rho$ & 0.971 & [0.961, 0.98] & -0.199 & [-0.377, -0.022]\\
\bottomrule
\end{tabular}}
\end{table}

How relevant is the difference between these different model
specifications? Figure \ref{fig:comparison-results} shows the
relationship between density and \(R_0\) implied by SWN Model 1 and
Selection Model 2. The left panel of the figure shows the non-linear but
monotonic relationship implied by SWN Model 1. The conclusion is that at
higher densities, \(R_0\) is \emph{always} higher. The right panel, in
contrast, shows that, according to Selection Model 2, \(R_0\) is zero
when density is zero (as expected), and then it tends to increase at
higher densities. This continues until a density of approximately 2.9
(1,000 people per sq.km). At higher densities than that \(R_0\) begins
to decline, and the relationship becomes negative at densities higher
than approximately 5.7 (1,000 people per sq.km).

Thus, other things being equal, the effect of density in a county like
Charlottesville in Virginia (density \textasciitilde1,639 people per
sq.km) is roughly the same as that in a county like Philadelphia
(density \textasciitilde4,127 people per sq.km). In contrast, the effect
of density on \(R_0\) in a county like Arlington in Virginia (density
\textasciitilde3,093 people per sq.km) is \emph{stronger} than either of
the previous two examples. Lastly, the density of counties like San
Francisco in California, or Queens and Bronx in NY, which are among the
densest in the US, contributes even less to \(R_0\) than even the most
rural counties in the country.

It is worth at this point to recall Cressie's dictum about modelling:
``{[}w{]}hat is one person's mean structure could be another person's
correlation structure'' {[}63{]}. There are almost always multiple ways
to approach a modelling situation. In the present case, we would argue
that spatial sampling is an important aspect of the modeling process,
but one that perhaps required different technical skills than those
available to SWN. There is nothing wrong with that. What matters is
that, by adopting relatively high reproducibility standards, these
researchers made a valuable and honest contribution to the collective
enterprise of seeking knowledge. Their effort, and subsequent efforts to
validate and expand on their work, can potentially contribute to provide
clarity to ongoing conversations about the relevance of density and the
spread of COVID-19.

In particular, it is noteworthy that a sample selection model with a
different variable transformation does not lend support to the thesis
that higher density is \emph{always} associated with a greater risk of
spread of the virus {[}as put by Wong and Li, ``\,`Density is destiny'
is probably an overstatement''; {[}46{]}{]}. At the same time, this also
stands in contrast to the findings of Hamidi et al., who found that
higher density was either not significantly associated with the rate of
the virus in a cross-sectional study {[}33{]}, or was negatively
associated with in a longitudinal setting {[}{[}51{]}. In this sense,
the conclusion that density does not aggravate the pandemic may have
been somewhat premature; instead, reanalysis of the data of SWN suggests
that Fielding-Miller et al. {[}44{]} might be onto something with
respect to the difference between rural and urban counties. More
generally, in population-level studies, density is indicative of
proximity, no doubt about that, but also for adaptive behavior. And it
is possible that the determining factor during COVID-19, at least in the
US, has been variations in perceptions of the risks associated with
contagion {[}31{]}, and subsequent compensations in behavior in more and
less dense regions.

\begin{figure}
\includegraphics[width=1\linewidth]{R0-Density-Reanalysis_files/figure-latex/comparison-results-1} \caption{\label{fig:comparison-results}Effect of density according to SWN Model 3 and Sample Selection Model 2.}\label{fig:comparison-results}
\end{figure}

@@

The tension between the need to publish research potentially useful in
dealing with a global pandemic, and a ``carnage of substandard
research'' {[}5{]}, highlights the importance of efforts to maintain the
quality of scientific outputs during COVID-19. An important part of
quality control is the ability of independent researchers to verify and
examine the results of materials published in the literature. As
previous research illustrates, reproducibility in scientific research
remains an important but elusive goal {[}e.g., 12,13--15{]}. This idea
is reinforced by the review conducted for this paper in the context of
research about population density and the spread of COVID-19.

Taking one recent example from the literature {[}Sy et al., {[}19{]};
SWN{]}, the present paper illustrates the importance of good
reproducibility practices. Sharing data and code can catalyze research,
by allowing independent verification of findings, as well as additional
research. After verifying the results of SWN, experiments with sample
selection models and variations in the definition of model inputs, lead
to an important reappraisal of the conclusion that high density is
associated with greater spread of the virus. Instead, the possibility of
a non-monotonical relationship between population density and contagion
is raised.

In the spirit of openness, this paper is prepared as an R Markdown
document, an a companion data package is provided. The data package
contains the relevant documentation of the data, and all data
pre-processing is fully documented. Hopefully this, and similar
reproducible papers, will continue to encourage others to adopt
reproducible standards in their research.

@@

\leavevmode\hypertarget{ref-Maddala1983limited}{}%
\CSLLeftMargin{62. }
\CSLRightInline{Maddala GS. Limited-dependent and qualitative variables
in econometrics. Cambridge: Cambridge University Press; 1983. }

\leavevmode\hypertarget{ref-Cressie1989geostatistics}{}%
\CSLLeftMargin{63. }
\CSLRightInline{Cressie N. Geostatistics. The American Statistician.
1989;43: 197.
doi:\href{https://doi.org/10.2307/2685361}{10.2307/2685361}}
